---
title: "CBSC185FinalProject"
author: "Naka Assoumatine"
date: "2024-12-12"
output: pdf_document
---

```{r Importing libraries, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(psych)
library(naniar)
library(lme4)
library(jtools)
```

# Import the class data 

```{r Import Data, include=FALSE}
class_data <- read.csv("Daily Survery Dec 3.csv")

```

```{r Create subset of the data,  include=FALSE}
class_data_subset <- subset(class_data, select = 
                          c(StartDate, ID, Q33, Q9_1))

# Delete the first and second rows; they are not data points
class_data_subset <- class_data_subset[-c(1,2), ]
```

```{r Initial summary of the data}
summary(class_data_subset)

glimpse(class_data_subset)
```

# Create a subset of the subdata with Id, Survey date mood, and hors of sleep

```{r Removing and renaming variables, include=FALSE}
# Remove all 100 becasue they were just placeholders
class_data_subset <- class_data_subset %>%
  filter(ID != "100")
class_data_subset <- rename(class_data_subset, 
                        survey_date = StartDate,
                        mood = Q33,
                        hours_of_sleep = Q9_1)
View(class_data_subset)
```

```{r Changing date format, include=FALSE}
class_data_subset$survey_date <- as.Date(sub(" \\d{2}:\\d{2}$", "", class_data_subset$survey_date), format = "%m/%d/%Y")

```

```{r Summary of the data after cleaning}
glimpse(class_data_subset)
```



```{r Change the values of hours_of_sleep to nummeric, include=FALSE}

# Initialize an empty vector for cleaned values
cleaned_sleep <- c()

# Loop directly over the values of hours_of_sleep
for (value in class_data_subset$hours_of_sleep) {
  
  # Case 1: Check if the value is numeric
  if (!is.na(as.numeric(value))) {
    cleaned_sleep <- c(cleaned_sleep, as.numeric(value))
  }
  
  # Case 2: Check if the value contains "hours" or "hour"
  else if ("hours" %in% unlist(strsplit(as.character(value), " ")) || 
           "hour" %in% unlist(strsplit(as.character(value), " "))) {
    numeric_part <- as.numeric(strsplit(as.character(value), " ")[[1]][1])
    cleaned_sleep <- c(cleaned_sleep, numeric_part)
  }
  
  # Case 3: Check if the value contains ":"
  else if (":" %in% unlist(strsplit(as.character(value), ""))) {
    parts <- unlist(strsplit(as.character(value), ":"))
    hours <- as.numeric(parts[1])
    minutes <- as.numeric(parts[2])
    cleaned_sleep <- c(cleaned_sleep, hours + minutes / 60)
  }
  
  # Default: Assign NA for invalid entries
  else {
    cleaned_sleep <- c(cleaned_sleep, NA)
  }
}

# Update the data frame with cleaned values
class_data_subset$hours_of_sleep <- cleaned_sleep
```

# Vizualize missing data
```{r Check missing data, echo=FALSE}
vis_miss(class_data_subset)
```

# Vizualize missing data after removing missing data
```{r Remove missing data and check again, echo=FALSE}
class_data_subset %>%
na.omit() %>%
  vis_miss()
```

#  Boxplot hours of sleep  
```{r Boxplot of Hours of sleep, echo=FALSE, warning=FALSE}
ggplot(class_data_subset, aes(x = "", y = hours_of_sleep)) +
  geom_boxplot(fill = "orange", color = "black", width = 0.2) +
  labs(title = "Boxplot of Hours of Sleep", 
       y = "Hours of Sleep", 
       x = "") +
  theme_minimal()
```


```{r Harcoding outlier, include=FALSE}
# Hardcoding for the case of 530
# I think the person probably meant 5:30 instead of 530
cleaned_sleep <- c()

for (value in class_data_subset$hours_of_sleep) {
  if (!is.na(value) && value == 530.0) {
    cleaned_sleep <- c(cleaned_sleep, 5.5)
  }
  else{
    cleaned_sleep <- c(cleaned_sleep, value)
  }
}
class_data_subset$hours_of_sleep <- cleaned_sleep
```


# Boxplot hours of sleep after harcoding the oulier 530
```{r Boxplot of Hours of sleep again, echo=FALSE, warning=FALSE}
# Boxplot of Hours of sleep again to check that there is no unrealistic hours of sleep
ggplot(class_data_subset, aes(x = "", y = hours_of_sleep)) +
  geom_boxplot(fill = "orange", color = "black", width = 0.2) +
  labs(title = "Boxplot of Hours of Sleep", 
       y = "Hours of Sleep", 
       x = "") +
  theme_minimal()
```
```{r change mood values 13, and 14, include=FALSE}
# Map the mood values to new values
class_data_subset$mood <- as.numeric(factor(class_data_subset$mood, 
                                            levels = c(1, 2, 3, 4, 5, 6, 7, 13, 14),
                                            labels = c(1, 2, 3, 4, 5, 6, 7, 8, 9)))
```



```{r Check missing data again, include=FALSE}
vis_miss(class_data_subset)
```

```{r Remove missing data and check again2, include=FALSE}
class_data_subset %>%
na.omit() %>%
  vis_miss()
```


# Histogram of mood
```{r Histogram of mood, echo=FALSE, warning=FALSE}
# Plot a histogram for mood
ggplot(class_data_subset, aes(x = mood)) +
  geom_histogram(binwidth = 1, fill = "orange", color = "black") +
  labs(title = "Histogram of Mood Responses", 
       x = "Mood Levels", 
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Bar plot of mood
To allow the reader to understand what the mood values are, I will create a bar plot with the values of the mood levels o the x-axis. This can be seen on the graph below:

```{r Creating a bar plot of mood, echo=FALSE, warning=FALSE}
#Factoring the values of the variable mood
class_data_subset$mood <- factor(class_data_subset$mood, 
                              levels = c(1:9), 
                              labels = c("extremely negative", 
                                         "mostly negative",
                                         "negative",
                                         "somewhat negative",
                                         "neither negative nor positive",
                                         "somewhat positive",
                                         "positive",
                                         "mostly positive",
                                         "extremely positive"))
ggplot(class_data_subset, aes(x = mood)) +
  geom_bar(binwidth = 1, fill = "orange", color = "black") +
  labs(title = "Bar plot of Mood Responses", 
       x = "Mood Levels", 
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```




```{r mood survey date for each individual over time, include=FALSE}
#Average sleep hours by ID
average_sleep <- class_data_subset %>%
  group_by(ID) %>%
  summarize(avg_sleep_hours = mean(hours_of_sleep))


# Ensure IDs are ordered correctly
class_data_subset$ID <- factor(class_data_subset$ID, levels = sort(as.numeric(unique(class_data_subset$ID))))

levels(class_data_subset$ID)
```


# Mood by Survey Date for Each Student
This will give an insight of the trend of the mood for all students

```{r plot each student, echo=FALSE, warning=FALSE}
# Compute average mood by survey_date
average_mood_over_time <- class_data_subset %>%
  group_by(survey_date) %>%
  summarize(avg_mood = mean(mood, na.rm = TRUE))  # Exclude missing values

ggplot(class_data_subset, aes(x = survey_date, y = mood, group = ID, color = ID)) +
  geom_line() +
  geom_point() +
  labs(title = "Mood by Survey Date for Each Student",
       x = "Survey Date",
       y = "Mood Level",
       color = "Student ID") +
  scale_x_date(breaks = average_mood_over_time$survey_date, 
               date_labels = "%b %d") +  # Format as "Month Day"
  theme_minimal() +
  theme(legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability


```

```{r revert mood back to numeric, include=FALSE}

# Create a mapping from labels back to numeric values
mood_mapping <- c("extremely negative" = 1, 
                  "mostly negative" = 2, 
                  "negative" = 3,
                  "somewhat negative" = 4,
                  "neither negative nor positive" = 5,
                  "somewhat positive" = 6,
                  "positive" = 7,
                  "mostly positive" = 8,
                  "extremely positive" = 9)

# Apply the mapping
class_data_subset$mood <- as.numeric(mood_mapping[class_data_subset$mood])


```


# Overall average mood overtime
```{r Plotting the average mood over time, echo=FALSE, warning=FALSE}
average_mood_over_time <- class_data_subset %>%
  group_by(survey_date) %>%
  summarize(avg_mood = mean(mood, na.rm = TRUE))  # Exclude missing values


ggplot(average_mood_over_time, aes(x = survey_date, y = avg_mood)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(title = "Average Mood Over Time", 
       x = "Survey Date", 
       y = "Average Mood") +
  scale_x_date(breaks = average_mood_over_time$survey_date, 
               date_labels = "%b %d") +  # Format as "Month Day" 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



# Regression model
```{r Run an LMER regression, echo=FALSE, warning=FALSE}
# Run the LMER model
lmer_model <- lmer(mood ~ hours_of_sleep + survey_date + (1 | ID), 
                   data = class_data_subset, 
                   REML = FALSE)

# Summary of the model
summary(lmer_model)
summ(lmer_model)
```

# Regression line of mood vs hours of sleep 

```{r Plot of predicted mood values, echo=FALSE, warning=FALSE}
complete_data <- na.omit(class_data_subset[, c("mood", "hours_of_sleep", "survey_date", "ID")])
complete_data$predicted_mood <- predict(lmer_model, re.form = NA)

class_data_subset <- merge(class_data_subset, 
                           complete_data[, c("ID", "survey_date", "predicted_mood")], 
                           by = c("ID", "survey_date"), 
                           all.x = TRUE)
ggplot(class_data_subset, aes(x = hours_of_sleep, y = mood)) +
  geom_point(alpha = 0.6, color = "blue") +  # Actual data points
  geom_line(aes(y = predicted_mood), color = "red", size = 1) +  # Regression line
  labs(title = "Regression Line of Mood vs Hours of Sleep",
       x = "Hours of Sleep",
       y = "Mood") +
  theme_minimal()


```




